Supplementary information: In-silico molecular enrichment and clearance of the human intracranial space

### 2025-02-27 paper/supplementary.tex
Split supplementary into separate file with independent references...

\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{bm}
\usepackage{caption}
\usepackage{subcaption}
% \captionsetup[subfigure]{justification=justified,singlelinecheck=false}
\usepackage{tikz}
\usepackage{newtxmath} % MER: Prettier math
\usepackage{cleveref} % MER: Allows \Cref (but use Section X.Y not Subsection X.Y)
\usepackage{booktabs} % MER: Prettier tables

@@

@@

\title{In-silico molecular enrichment and clearance of the human intracranial space}
\author[1]{Marius Causemann}
\author[1]{Miroslav Kuchta}
\author[2]{Rami Masri}
\author[1,3,*]{Marie E. Rognes }
\affil[1]{Department of Numerical Analysis and Scientific Computing, Simula Research Laboratory, Oslo, Norway}
\affil[2]{Division of Applied Mathematics, Brown University, Providence, Rhode Island, USA}
\affil[3]{K. G. Jebsen Centre for Brain Fluid Research, University of Oslo, Norway}
\affil[*]{meg@simula.no}

\newcommand{\rami}[1]{\textcolor{blue}{#1}}
\newcommand{\mer}[1]{\textcolor{magenta}{#1}}
\newcommand{\mar}[1]{\textcolor{violet}{#1}}
\newcommand{\mk}[1]{\textcolor{orange}{#1}}
\newcommand{\discuss}[1]{\textcolor{red}{#1}}
\newcommand{\fixme}[1]{\textcolor{red}{#1}}
\newcommand{\draft}[1]{\textcolor{lightgray}{#1}}

@@

\begin{abstract}
  The mechanisms of intracranial solute transport are fundamental to human brain health, with alterations often linked to disease and functional impairment, and with distinct opportunities for personalized diagnostics and treatment. However, our understanding of these mechanisms and their interplay remains incomplete, in part due to the complexity of integrating insights across scales, between species and from different modalities. Here, we combine mixed-dimensional modelling, multi-modal magnetic resonance images, and high performance computing to construct and explore a high-fidelity in-silico model of human intracranial molecular enrichment.  This model predicts the temporo-spatial spreading of a solute within an image-derived geometric representation of the subarachnoid space, ventricular system and brain parenchyma, including networks of surface perivascular spaces (PVSs). Our findings highlight the significant impact of cerebrospinal fluid (CSF) production and intracranial pulsatility on molecular enrichment following intrathecal tracer injection. We demonstrate that low-frequency vasomotion induces moderate CSF flow in surface PVS networks which substantially enhances tracer enrichment, and that impaired enrichment is a direct natural consequence of enlarged PVSs. This openly available technology platform thus provides an opportunity for integrating separate observations on diffusion in neuropil, vascular dynamics, intracranial pulsatility, CSF production, and efflux, and for exploring drug delivery and clearance in the human brain.
\end{abstract}
\begin{document}

\thispagestyle{empty}

%\mer{MER: Target length: 5000 words (Intro, Results, Discussion) plus <3000 words Methods: 700 words Introduction. 3100 words Results. 1200 words Discussion.}

\section*{Introduction}

% What is the big picture problem and why is it important? 
The mechanisms underlying molecular transport within the intracranial
space are fundamental to human brain health and
function. Neurodegenerative diseases such as Alzheimer's, Parkinson's,
and Huntington's disease are all associated with abnormal accumulation
of protein aggregates together with alterations in transport and
clearance characteristics~\cite{rasmussen2018glymphatic,
  harrison2020impaired, eide2023plasma, liu2024glymphatic}. Moreover,
sleep and conversely sleep-deprivation play a definite yet enigmatic
role in modulating molecular enrichment and
clearance~\cite{xie2013sleep, eide2021sleep, eide2022altered,
  miao2024brain, hauglund2025norepinephrine}. In the last decade,
established theories have been challenged by new findings revealing a
greater degree of molecular movement and
exchange~\cite{iliff2012paravascular, ringstad2017glymphatic,
  louveau2017understanding, proulx2021cerebrospinal,
  bohr2022glymphatic}, as well as substantial variability in
enrichment and clearance between individuals and between patient
cohorts~\cite{ringstad2018brain, eide2021direction, eide2021impaired,
  eide2022altered}. These observations provide distinct opportunities
(and challenges) for personalized medicine e.g.~for tailored
intrathecal delivery of chemotherapy~\cite{lohela2022glymphatic} and
for early diagnostics of impaired brain
clearance~\cite{eide2021clinical, van2024human}. In spite of their
importance, our understanding of these mechanisms is incomplete with
open challenges and significant debate -- in part relating to the
translation of knowledge between scales, species, experimental
protocols, clinical cohorts, and individuals.

Perivascular pathways along the brain surface and within the brain
parenchyma have long been hypothesized to serve a designated role in
this context~\cite{rennels1985evidence, zhang1990interrelationships,
  ichimura1991distribution, carare2008solutes, iliff2012paravascular,
  foley2012realtime, hannocks2018molecular, van2024caa}. Recently, Eide and
Ringstad~\cite{eide2024functional} and Yamamoto et
al~\cite{yamamoto2024perivascular} demonstrated that perivascular
spaces (PVSs) define preferential pathways for molecular transport in
humans, with delayed periarterial enrichment in dementia
subtypes~\cite{eide2024functional}. Perivascular flow of cerebrospinal
fluid (CSF) clearly contributes to this transport, and is inherently
associated with vascular pulsations~\cite{hadaczek2006perivascular,
  iliff2013cerebral, bedussi2018paravascular, mestre2018flow,
  boster2023artificial, hirschler2024region}. Fluid mechanics
considerations point at intracranial pressure differences and shorter
wavelength vascular wall pulsations as drivers of directional net flow
and convection in the PVS~\cite{bilston2003arterial, rey2018pulsatile,
  daversin2020mechanisms, kedarasetti2020functional,
  gjerde2023directional, nozaleda2024arterial}, while longer waves
such as the pulse wave primarily contribute to oscillatory flow and
dispersion~\cite{asgari2016glymphatic, sharp2019dispersion, thomas2019fluid,
  kedarasetti2020arterial, troyetsky2021dispersion, 
  martinac2021phase}. In the bigger picture, CSF is produced by the
choroid plexus~\cite{damkier2013cerebrospinal,
  steffensen2018cotransporter, liu2020direct}, pulsates through the
ventricular system, cisterns, and subarachnoid space (SAS) in
synchrony with cardiac, respiratory, and neural
waves~\cite{greitz1993pulsatile, wagshul2011pulsating,
  sweetman2011cerebrospinal, fultz2019coupled, vinje2019respiratory,
  eide2021direction, causemann2022human, williams2023neural,
  zimmermann2023total}, and drains via the dural sinuses, meningeal
lymphatics, cranial nerves, or other efflux
pathways~\cite{proulx2021cerebrospinal}. However, how these
physiological factors and physical mechanisms integrate to enhance or
impair human intracranial molecular transport over larger spatial
scales and longer time scales remain unknown. A related key question
is to what extent surface PVSs are separated from the SAS by
structural barriers~\cite{zhang1990interrelationships,
  weller2005microscopic, bedussi2017paravascular,
  pizzo2018intrathecal, mestre2022periarteriolar,
  mollgard2023mesothelium, smets2024perivascular, eide2024functional},
and in turn to what extent such structural compartmentalization is a
prerequisite for effective perivascular transport.

% What are our research questions and how do we answer them
In this study, by leveraging geometric model reduction and
mixed-dimensional modelling~\cite{masri2024modelling}, structural
magnetic resonance (MR) images~\cite{hodneland2019new}, and high
performance computing, we introduce an integrated computational model
of intracranial molecular enrichment and clearance. Focusing on the
interplay between perivascular pathways, pulsatility and CSF flow
dynamics, the model predicts the temporal evolution and spatial
distribution of a solute concentration within a detailed geometric
representation of the human SAS and ventricular system, networks of
surface PVSs and the brain parenchyma. In terms of transport dynamics,
we account for heterogeneous diffusion, dispersive mixing induced by
cardiac and respiratory pulsatility in the CSF spaces and PVSs,
convective fluid flow driven by CSF production and peristaltic
pumping, as well as solute exchange and clearance across
semi-permeable membranes. Comparing with glymphatic MRI
studies~\cite{ringstad2017glymphatic, ringstad2018brain,
  watts2019measuring}, the in-silico predictions accurately represent
molecular enrichment patterns, timing and intercompartmental
distributions. This open platform~\cite{ZENODO} thus provides a
technological opportunity for qualitatively and quantitatively
exploring key open questions relating to molecular movement within the
human brain environment such as the role of pulsatility, perivascular
pathways, structural compartmentalization, or morphology.

% Main findings and outlook
By exploring this high-dimensional parameter space, we propose that
the balance between CSF production and intracranial pulsatility is key
to shaping the large-scale features of intracranial enrichment
patterns, with the potential to span a wide range of individual and
cohort variability. Moreover, we predict that CSF production, cardiac-
and respiratory pulsatility is not sufficient to explain early
perivascular enrichment, but that fluid flow induced by low-frequency
vasomotion in surface periarterial spaces (on the order of 10 \textmu
m/s) is sufficient, even in the absence of a structural
compartmentalization of the PVS. Conversely, enlarged PVSs in the SAS
will cause a substantial reduction in cardiac- and vasomotion-driven
flow velocities, strongly delay perivascular transport, and thus
impair intracranial enrichment. These findings transfer, reconcile,
integrate and extend insights from clinical, experimental, and
theoretical studies, and lay a new foundation for in-silico studies of
personalized intrathecal drug delivery and brain clearance.

\section*{Results}

\subsection*{In-silico predictions of intracranial molecular enrichment and clearance after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/figure1.png}
     \caption{
     \textbf{Our in-silico model predicts tracer transport in PVS, SAS and parenchyma following intrathecal injection.}\\
     A) Illustration of the spatial domains of the model: CSF-filled space (ventricles and SAS) $\Omega_{\rm CSF}$, parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and the veins $\Lambda_{\rm vein}$ and their interfaces and boundaries (see Methods); 
     B) 3D rendering of the actual geometry used in this study parenchyma, CSF (clipped, blue), arterial network (red), venous network (blue)); 
     C) illustration of the modelled clinical tracer injection procedure; 
     D) simulated tracer concentration after 1, 6, 12, and 24 hours (opacity increases linearly from 0 to $2\,$mol/l); 
     E) sagittal, coronal and axial view of computed tracer concentrations over T1-image after 1, 6, 12 and 24\,h (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); 
     F) average tracer concentration in each compartment over the first 24\,h. 
     G) total amount of tracer in each cranial compartment over the first 24\,h; }
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--B). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}. Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

Simulating a glymphatic MRI protocol~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2024functional}(\Cref{fig:results1}C), we then predict the
spreading of $0.5$ mmol intrathecally injected Gadobutrol after its
appearance at the craniocervical junction
(\Cref{fig:results1}D--E). After one hour, the in-silico
tracer moves upwards in the SAS frontally of the brainstem, and
quickly reaches the supratentorial regions. Here, it spreads both
posterior through the quadrigeminal cistern and the longitudinal
fissure, and anteriorly through the outer SAS, reaching the top of the
cerebral cortex after around 12 hours. After 24 hours, the tracer
covers most of the brain surface (with the exception of some posterior
regions) and has penetrated substantially into the parenchymal
tissue. This pattern is reflected in the mean tracer concentrations in
each compartment, where the CSF space reaches its peak of
$1.4\,$mmol/l after 2 hours, followed by the arterial PVS
concentration peak ($2.9\,$mmol/l after 4 hours)
(\Cref{fig:results1}F). While the mean concentrations in these
compartments drop soon after peaking, the parenchymal tissue
slowly enriches with tracer over the first 24 hours, with a final
value of $0.3$ mmol/l. Overall, about 40\% of the total amount of
tracer remains in the cranium 24 hours post-injection, with the largest
share in the CSF (53\%), followed by the parenchyma (35\%), the
venous PVS (8\%) and the arterial PVS
(5\%) (\Cref{fig:results1}G).

\subsection*{Reduced CSF pulsatility strongly shifts intracranial enrichment patterns}
\begin{figure}%[h!]
\centering 
\includegraphics[width = 0.95\textwidth]{figures/figure2.png}
\caption{\textbf{The balance between CSF pulsatility and production shapes molecular enrichment patterns.}
A) CSF spaces: the third and fourth ventricle (V3\&4), lateral ventricles (LV) and SAS; 
B) Boundaries and interfaces: the spinal subarachnoid space (SSAS), upper and lower arachnoid membrane (AM-U and AM-L), pia (Pia) and lateral ventricles (LV); 
C) Schematic of the CSF flow model; 
D) CSF flow induced by CSF production with outflow allowed across the
upper convexity (AM-U); 
}
\label{fig:csf}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
E) Histograms of the production-induced CSF velocities in the SAS and third and fourth ventricle (V3 \& V4) in terms of the relative frequency of the mean flow speed in each computational cell weighted by its volume; 
F) CSF flow induced at peak systolic blood inflow; 
G) Cardiac-induced dispersion factor $R_c$;
H) CSF flow induced at peak respiratory expansion; 
I) Respiration-induced dispersion factor $R_r$, $D = (1 + R_c + R_r) D^{\rm Gad}$; J--M): Histograms of the cardiac-induced flow velocities (J),
cardiac-induced dispersion factors (K),
respiratory flow velocities (L), respiratory dispersion factor (M);
N--Q): sagittal, coronal and axial planes of tracer concentration after 12h for different model variations: baseline (N), no CSF production (O), low dispersion (P), and high dispersion (Q).
}
\end{figure}

The enrichment patterns observed clinically after intrathecal
injection of contrast differ substantially between subjects and
between pathological conditions~\cite{ringstad2018brain,
  eide2021direction, eide2021impaired, eide2022altered}. These
neurological conditions are also associated with alterations in the
pulsatile flow of CSF in the ventricular system and
SAS~\cite{eide2021direction}. Clearly, key physiological factors such
as cardiac pulsatility and respiration easily differ between
individuals and between both pathological and physiological states. We
therefore next asked to what extent -- and how -- variations in CSF
pulsatility would affect the intracranial enrichment characteristics.

For the CSF dynamics in our baseline model, we combine the net
contribution to CSF flow induced by CSF production with the integrated
dispersive effects of cardiac and respiratory pulsatile CSF flow
(\Cref{fig:csf}A--C). Imposing a constant CSF production of
$400\,$ml/day across the surface of the lateral ventricles, while
allowing for efflux across the upper convexity, yields a total CSF
pressure drop of $26\,$mPa (0.00020 mmHg) (\Cref{fig:csf}D) with a
maximum flow velocity of $1.85\,$mm/s (in the aqueduct) and a mean
velocity in the SAS of $2.6$\textmu m/s (\Cref{fig:csf}E). To examine the
dispersive effects of pulsatile CSF flow, we employ the theory of
shear-augmented dispersion together with computational fluid dynamics
to determine a dispersion factor $R$ enhancing the effective solute
diffusivity (see Methods). For the cardiac contribution, we compute
CSF pressure and flow fields at peak systolic blood inflow,
corresponding to a reduction in CSF space volume at a total rate of
$6\,$ml/s~\cite{baledent2014imaging, causemann2022human} in the SAS
and $0.31\,$ml/s across the lateral ventricle
surface\cite{vinje2019respiratory}. This scenario sets up a pressure
drop of $10\,$Pa ($0.075$ mmHg) between the lateral ventricles and the
spinal SAS and a maximum flow velocity of $19.8\,$cm/s
(\Cref{fig:csf}F, J). Assuming a cardiac frequency of $1\,$Hz, we
infer that this cardiac-induced pulsatile CSF flow increases the
effective diffusion by more than two orders of magnitude in the
aqueduct and near the cisterna magna, but has little effect ($R <
1$) in most of the SAS (\Cref{fig:csf}G, K). For the respiratory
contribution, we employ the same methodology, but with a total rate of
$1\,$ml/s \cite{gutierrez2022effect} in the SAS and $0.121\,$ml/s
\cite{liu2024using} in the lateral ventricles, yielding a respiratory
peak flow volume of $1.121\,$ ml/s at the craniocervical
junction. While the resulting flow velocities are only about one
fourth of their cardiac-induced counterparts (\Cref{fig:csf}H, L),
respiratory dispersive mixing reaches a factor of up to 320 due the
lower respiratory frequency of $0.25\,$Hz (\Cref{fig:csf}I, M).

Now, to examine how CSF pulsatility affects the enrichment patterns,
we consider three variations of the baseline: (i) no CSF production,
(ii) reduced pulsatility and thus decreased dispersion ($0.1 \times
R$), and (iii) higher pulsatility with increased dispersion ($10
\times R$). Without CSF production, transport is considerably delayed
(\Cref{fig:csf}N, O). Even after $12$ hours, tracer remains in the
subtentorial regions around the cerebellum, the brain stem and in the
surrounding CSF. Interestingly, the lack of CSF production instead
allows tracer to travel upwards through the ventricular system,
reaching the third ventricle after around $12$ hours. On the other
hand, if the CSF pulsatility is reduced, we observe rapid transport
towards the upper convexity of the cranium, as in the baseline model,
but the tracer spreads exclusively within the anterior regions
(\Cref{fig:csf}P). This feature can be attributed to the CSF flow
bifurcation posterior to the ambient cistern (\Cref{fig:csf}D):
without sufficient diffusion, the tracer is unable to cross into the
posterior SAS. Indeed, with higher dispersion, the tracer moves
through the quadrigeminal cistern and further upwards into the
longitudinal fissure, with also enrichment of the cerebellum
(\Cref{fig:csf}Q).

\subsection*{Perivascular flow shapes and accelerates molecular enrichment}
\label{sec:pvs_flow_results}

The PVSs are recognized across species as critical pathways for solute
transport in and around the brain, and thus as potential targets for
enhancing brain drug delivery and metabolic waste clearance. However,
whether CSF flows more rapidly in PVSs and what the forces and
mechanisms required to drive such flow are, remain as key points of
debate~\cite{bohr2022glymphatic, van2024caa}. Motivated by
experimental observations in animal
models~\cite{iliff2012paravascular, iliff2013cerebral, mestre2018flow,
  bedussi2018paravascular}, there is now a remarkable body of
literature on modelling perivascular fluid flow and
transport~\cite{bilston2003arterial, asgari2016glymphatic,
  rey2018pulsatile, daversin2020mechanisms, sharp2019dispersion,
  thomas2019fluid, kedarasetti2020functional, kedarasetti2020arterial,
  troyetsky2021dispersion, martinac2021phase, gjerde2023directional,
  nozaleda2024arterial}. Here, we ask how these proposed mechanisms
would translate from idealized geometries to human vascular networks
and moreover, evaluate their integrated effect in the context of
intracranial solute transport.

Our periarterial network extends from the internal carotid arteries
and basilar artery through up to 18 bifurcations to reach upstream
network ends located within the SAS or up to 6 mm inside the
parenchyma (\Cref{fig:results1}A--B, \Cref{fig:pvs_a}A) and includes
major surface arteries of radius 0.5--1.4 mm. Imposing the pressure
field induced by CSF production at the periarterial network ends
induces slow steady CSF flow of variable direction in these PVSs, with
an average velocity of 0.08 \textmu m/s (antegrade) and a maximum
velocity of 0.55 \textmu m/s (\Cref{fig:pvs_a}B). On the other hand,
traveling waves of arterial wall motion (\Cref{fig:pvs_a}C), such as
the pulse wave or other vasomotion~\cite{vanveluw2020vasomotion,
  munting2023spontaneous, bojarskaite2023sleep,
  broggini2024long,hauglund2025norepinephrine}, also induce net
directional flow in the PVS~\cite{kedarasetti2020arterial,
  kedarasetti2020functional, coenen2021lubrication,
  gjerde2023directional, nozaleda2024arterial} -- of magnitude and
direction depending on the amplitude, frequency, and length of the
waves and the characteristics of the perivascular
network~\cite{gjerde2023directional}. Applying a semi-analytic model
of the net flow induced by peristalsis in perivascular
networks~\cite{gjerde2023directional} (see Methods), we estimate that
the cardiac pulse wave alone, traveling at a frequency of 1 Hz with a
wavelength of 2.0 m and a 1\% wall displacement~\cite{jung2021novel},
will induce mainly antegrade PVS flow with an average net velocity of
0.92 \textmu m/s while reaching up to 7.31 \textmu m/s near larger,
ventral vessels such as the MCA (\Cref{fig:pvs_a}D,E). The same theory
predicts that strong ultraslow vasomotions, if traveling antegrade at
0.1Hz with a wavelength of 0.02 m and a 10\%
displacement~\cite{broggini2024long}, will induce both retrograde and
antegrade net PVS flow with an average velocity of 13.05 \textmu m/s
and maximum velocity 54.44 \textmu m/s (\Cref{fig:pvs_a}D,F).

We study the effect of such rapid PVS flow on intracranial molecular
enrichment by comparing the more conservative baseline model and a
high PVS flow model, where the former still includes the net flow
contributions from CSF production and the cardiac pulse wave, while
the latter additionally includes the ultraslow vasomotion
contribution. In both models, tracers were first observed at the basal
artery after 48 minutes (\Cref{fig:pvs_b}A), defined by their first-time arrival (FTA), the time at which the concentration first exceeds $0.1$ mmol/l. At all upstream locations along the middle cerebral arteries
(MCAs) and anterior cerebral artery (ACA), we found substantially
reduced FTAs with higher PVS flow (\Cref{fig:pvs_b}B), up to 2.5 hours
earlier in the left M2 segment of the MCA. PVS concentrations peaked
before the surrounding tissue (0:48h vs 1:12h at the MCA and 1:00h vs
1:48h at MCA2) with high PVS flow, while these peaks occurred nearly
simultaneously in the baseline model. The earlier appearance and the
delay in time-to-peak between the PVS and surrounding tissue clearly
indicates the directionality of tracer enrichment -- it first arrives
in the PVS and subsequently spreads into the tissue and SAS
(\Cref{fig:pvs_b}C, D), and especially so with higher PVS flow. These
observations also hold on an aggregated level: computing the total
amount of tracer in the PVS as a function of the distance to the
arterial network roots, we find accelerated tracer transport with
higher PVS flow, especially at earlier time points (2--9 hours)
(\Cref{fig:pvs_b}E). Moreover, differences in tracer transport along the
PVS translate to altered enrichment patterns on the whole-organ
scale. For instance, after 4-6 hours, the faster-moving tracer in the
PVS is clearly visible in the space adjacent to the PVS of the ACA, MCA, and other arteries (\Cref{fig:pvs_b}G). In contrast,
there are no clear signs of early enrichment surrounding the PVS in
the baseline model (\Cref{fig:pvs_b}F).

A natural question is whether early PVS enrichment could be the result
of increased dispersion rather than net flow in the
PVS\cite{asgari2016glymphatic,sharp2019dispersion,bojarskaite2023sleep,asgari2016glymphatic,troyetsky2021dispersion}. Interestingly,
even increasing the dispersion factor by $100 \times$ in the PVS,
induced only minor changes in the global spreading rate compared to the
baseline (\Cref{fig:pvs_b}F, H), thus indicating a negative answer to
this question.

\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure3a.png}
    \caption{\textbf{A static pressure gradient and peristaltic pumping are potential drivers of perivascular flow.}
    A) Surface arterial network colored by arterial radius with vessel locations labeled;
    B) Net PVS flow induced by CSF production (cf.~\Cref{fig:csf}
    -- histograms show the relative frequencies of local periarterial velocities. Positive values indicate antegrade PVS flow, while negative values indicate retrograde PVS flow;
    C) Illustration of the concept of peristaltic pumping;
    D) histograms of estimated net PVS flow induced by pulse wave or vasomotion peristaltic pumping;
    E) Estimated net PVS flow induced by pulse wave peristaltic pumping.
    F) Estimated net PVS flow induced by vasomotion/slow wave peristaltic pumping.}
    \label{fig:pvs_a}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.99\textwidth]{figures/figure3b.png}
    \caption{\textbf{Perivascular flow shapes and accelerates molecular enrichment.}
    A) Comparison of mean tracer concentration in the PVS and over the outer PVS surface at key locations of the arterial tree, for the baseline (lower) and high PVS flow (upper) models, with labels: ``peak'': time-to-peak (h), ``$\Delta t$'': time difference between peaks (h), and ``FTA'': first-time tracer arrival (h).
    B) first-time arrival for the main trunks of the periarterial network over the distance from the closest network root node (BA, ICA-R or ICA-L for the baseline, the high PVS flow and the high PVS dispersion models);}
    \label{fig:pvs_b}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
    C, D) 2D slices showing zoom-in on the region surrounding the MCA-R and MCA2-R for the baseline (upper) and high PVS flow (lower) models. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF));  
    E) Total amount of tracer in the periarterial spaces as a function of the distance to the nearest network root over time for the baseline (left) and the high PVS flow (right) models;
    F--H) sagittal (upper) and coronal (lower) views of the simulated tracer concentrations overlayed on the T1-weighted MR image after 4 and 6h for the baseline (K), high PVS flow (L) and high PVS dispersion (M) models (concentration opacity linearly increasing with concentration value, pial surface outlined in pink, arachnoid membrane in cyan, and arteries in dark red). An interactive visualization of the baseline and the high PVS flow model can be found \href{https://baseline-pvs-flow.surge.sh/}{here}, and \href{https://high-pvs-flow.surge.sh/}{here}.}
\end{figure}

\subsection*{Structural versus functional compartmentalization of perivascular spaces}

Human and rodent observations indicate that tracers concentrate in perivascular spaces surrounding the pial and subarachnoid vasculature \cite{zhang1990interrelationships,zhang1992directional, bedussi2017paravascular, mestre2018flow, bedussi2018paravascular, eide2024functional}. However, it remains unclear whether such enrichment patterns necessitate a structural barrier, such as a membrane with limited permeability, or if the patterns could result from enhanced flow or mixing in these areas alone. We therefore next investigate how the permeability of the interface between the PVS and the surrounding CSF affects tracer enrichment around the major arterial trunks (\Cref{fig:compartmentalization}A--B). To this end, we compare models with high and low  permeabilities, representing a highly permeable PVS-CSF interface (functional compartmentalization) and a less permeable interface (structural compartmentalization), respectively. For the low permeability model, we set the permeability to $3.8 \cdot 10^{-7}\,$m/s, consistent with previous estimates for the endfoot sheath surrounding penetrating arterioles \cite{koch2023estimates}, which we consider to be a lower bound for the surface PVS-CSF permeability. For the high permeability model, we increase this permeability by a factor of 100. For both scenarios, we consider the high PVS flow regime as examined in the previous section.

At the basal artery (BA), tracer appears within one hour in the surrounding CSF in both models (\Cref{fig:compartmentalization}C, E). With a more permeable PVS-CSF interface, tracer quickly crosses from the CSF into the PVS, resulting in similar peak concentrations of 25\,mmol/l in the PVS and at its outer surface, while values up to 50\,mmol/l are attained in the vicinity. In contrast, the low permeability model exhibits substantially lower perivascular tracer enrichment, with a maximum concentration of 10\,mmol/l after approximately two hours.
%The high permeability model allows for more tracer crossing into the PVS, leading to higher concentrations further downstream. Notably, crossover into the parenchyma is also observed, most clearly at 4 hours.
For the middle cerebral arteries (MCA-R and MCA-L), tracer appears
first in the PVS in both models (\Cref{fig:compartmentalization}D, E),
though with higher concentrations in the high permeability model (up
to 24 mmol/l). We also observe that the higher permeability allows the
tracer to leak out of the PVS and spread within the Sylvian fissure to
a greater extent. With a small delay, more tracer appears via
  the CSF pathway (\Cref{fig:compartmentalization}D). At the anterior cerebral artery (ACA-A2), the first tracer arrival in the PVS occurs after 3 hours, with a peak at 5 hours. The concentration peak in the PVS is tailed by a higher peak in the CSF (\Cref{fig:compartmentalization}E), indicating tracer arrival through a second pathway; i.e., directly through the CSF-filled space outside of the PVS. Similar patterns are observed for ACA-A3.

Summarizing and quantifying these observations
(\Cref{fig:compartmentalization}F), we find that neither the time of
first arrival nor the time-to-peak are substantially affected by the
PVS-CSF permeability at any of the periarterial segments considered
(BA, MCAs, ACAs). However, the concentration difference between the
PVS and its surrounding CSF quickly decreases with higher
permeability, as does the time lag between the concentration peaks in
these domains. Thus, fast transport along the PVS is not contingent
upon a structural compartmentalization of the PVS, whereas a sharp
concentration gradient between PVS and CSF is unlikely without a
restricting barrier.
\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure4.png}
    \caption{\textbf{Does early perivascular enrichment rely on a structural compartmentalization of the PVS?}
    A) The PVS-CSF membrane permeability regulates the exchange of solutes between the PVS and surrounding CSF spaces; 
    B) The arterial network with highlighted regions of interest; 
    C) 2D slices showing zoom-in to the region surrounding the basal artery after 1, 2, 3, and 4 hours for a low (above) and high (below) PVS-CSF permeability model. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF));
    D) As in (C) for regions surrounding (from upper to lower) the ACA-A3, ACA-A2, MCA-R and MCA-L segments;}
    \label{fig:compartmentalization}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
E) Comparison of mean tracer concentration in the PVS (dashed) and over the
outer PVS surface (dotted) for the BA, MCA-R, MCA-L, ACA-2, and ACA3 segments over time for the high (purple) and low permeability (yellow) models;
F) Effect of varying the PVS-CSF permeability on first-time-of-arrival (FTA), time difference ($\Delta t$) between concentration peaks in the PVS and surrounding tissue, time of PVS peak concentration, PVS peak concentration and difference in peak concentration between PVS and surrounding tissue.}
\end{figure}

\subsection*{Enlarged PVSs delay periarterial and intracranial molecular enrichment}

Enlarged PVSs are associated with impaired periarterial tracer
transport and more generally with cognitive decline as well as
certain subtypes of
dementia~\cite{bown2022physiology,eide2024functional}. A key
question is thus whether enlarged PVSs alone leads to delayed perivascular
and intracranial tracer enrichment. To address this question from a
fluid and transport dynamics perspective, we ask our in-silico model
to predict the integrated effect of perivascular dilation on PVS flow
velocities and tracer enrichment patterns (\Cref{fig:enlargement}A).

>>>>

\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{bm}
\usepackage{subcaption}
% \captionsetup[subfigure]{justification=justified,singlelinecheck=false}
\usepackage{tikz}
\usepackage{newtxmath} % MER: Prettier math
\usepackage{cleveref} % MER: Allows \Cref (but use Section X.Y not Subsection X.Y)
\usepackage{booktabs} % MER: Prettier tables

@@

\renewcommand\thefigure{S\arabic{figure}}    
\renewcommand\thesection{S\arabic{section}}

@@

\title{Supplementary information: In-silico molecular enrichment and clearance of the human intracranial space}
\author[1]{Marius Causemann}
\author[1]{Miroslav Kuchta}
\author[2]{Rami Masri}
\author[1,3,*]{Marie E. Rognes }
\affil[1]{Department of Numerical Analysis and Scientific Computing, Simula Research Laboratory, Oslo, Norway}
\affil[2]{Division of Applied Mathematics, Brown University, Providence, Rhode Island, USA}
\affil[3]{K. G. Jebsen Centre for Brain Fluid Research, University of Oslo, Norway}
\affil[*]{meg@simula.no}

\begin{abstract}
  Supplementary methods and discussion accomanying the main text of
  \emph{In-silico molecular enrichment and clearance of the human
  intracranial space}.
\end{abstract}

\begin{document}

@@

\input{supplementary_main.tex}
